Towards a standardized evaluation of imputation routines

### 2022-03-22 Manuscript/StandardizedEvaluation.Rmd
Manuscript after discussion on suggestions Hanne, including first version of the checklist.

---
title: "Towards a standardized evaluation of multiple imputation routines"
author: "Hanne Oberman and Gerko Vink"
date: "19-1-2022"
abstract: "Developing new imputation methodology has become a very active field. Unfortunately, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. In this paper we propose a move towards a standardized evaluation of imputation methods. To demonstrate the need for standardization, we highlight a set of potential pitfalls that bring forth a chain of potential problems in the objective assessment of the performance of imputation routines. This may lead to suboptimal use of multiple imputation in practice. Additionally, we suggest a course of action for simulating and evaluating missing data problems."
bibliography: StandardizedEvaluation.bib  
output: 
  pdf_document:
    keep_tex: true
editor_options: 
  chunk_output_type: inline
---

@@

<!-- # Structure -->

<!-- 1. introduction -->

<!-- - aim -->
<!-- - number of simulations -->

<!-- 2. data generation -->

<!-- - model vs design based -->
<!-- - data types -->
<!-- - sample vs population (vs bootstrapping??) -->
<!-- - number of observations -->
<!-- - number of variables -->
<!-- - coherence between variables -->

<!-- 3. missingness generation -->

<!-- - monotone/univariate vs multivariate -->
<!-- - missingness mechanism -->
<!-- - type of M(N)AR -->
<!-- - missingness proportion (vs proportion of incomplete cases! 50% incomplete == 25% missing in bivariate data) -->

<!-- 4. missing data methods -->

<!-- - imputation method -->
<!-- - number of imputations -->
<!-- - number of iterations -->

<!-- 5. performance evaluation -->

<!-- - estimand -->
<!-- - diagnostic -->
<!-- - ppc -->

@@

Multiple imputation [@rubin1987] is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policy making, demonstrated e.g. by the detailed manual created by the National Research Council [@little2012prevention]. Although a top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers are embracing multiple imputation techniques. After all, the principle of multiple imputation is very intuitive.

The idea behind multiple imputation is to impute (fill in) the missing values in incomplete data multiple times, to obtain a valid estimate of what could have been observed. The multiple datasets that are thus obtained can be analyzed by standard techniques and the analysis results can be combined into a single inference. In contrast to ad hoc methods for dealing with missing values (e.g. list-wise deletion, mean imputation, last observation carried forward), multiple imputation properly takes the sources of uncertainty that are related to the missingness problem into account. [It has been empirically demonstrated time and again that the inferences on multiply imputed data yield unbiased and confidence valid results (through Rubin's rules). But how do we know that this is the case, if we do not have access to the true but unobserved values of the missing entries? And how can we be sure that extensions to the methodology yield valid inferences too? Especially in the field of ML/AI, there are promising new imputation methods, which may yield even sharper imputations than now-standard (semi-)parametric imputation methods. That's why we use simulation studies for the evaluation of imputation routines.]

~~The quality of a solution obtained by multiple imputation depends on the statistical properties of the incomplete data and the degree to which an imputation procedure is able to capture these properties when modeling missing values. In general it holds that modeling missing data becomes more challenging when the amount of missingness increases. However, when (strong) relations in the data are present, the observed parts can hold great predictive power for the models that estimate the missingness. In that case, multiple imputation would be substantially more efficient than the ubiquitous list-wise deletion.~~

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. In such studies a complete dataset is usually generated from a statistical model, another model is used to induce missingness, [then several missing data methods are applied] and a set of evaluation criteria is postulated to evaluate the performance of the imputation methods. However, no golden standard has been established to evaluate these imputation routines and, as a result, the validity of the simulation procedures may differ tremendously from one developer to another.

@@

To evaluate the ability of an imputation routine to handle ~~the non-response model~~ [missing data entries], a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand, by choosing a data generating mechanism to obtain a complete dataset.  Complete data is typically either generated from a statistical model, or sampled from a sufficiently large ~~observed~~ set [model-based vs design-based simulation]. ~~However, when generating data, two easily overlooked problems may arise.~~

@@

@@

[Briefly introduce missingness mechanisms here and show potential problems, but without simulation results (moved to separate section). This part should contain: 1) definitions of mechanisms, 2) that MCAR is often ignored, while it's the minimum requirement and sometimes realistic in practice, 3) that MAR can be generated *spuriously* if the correlation between variables is low, and 4) that MNAR is often ignored, while it may be the most realistic in practice. And finally something about types of M(N)AR, but what are specific problems with that?]

@@

~~However, because multiple imputation was designed as a mode for statistical inference, a minimal set of statistical properties should at least be evaluated. We highlight four aspects of evaluation that should be considered when evaluation imputation routines.~~

@@

@@

# Illustration

[Add simulation study + results here!]

@@

## Step 3. Evaluate performance
It is wise to evaluate the performance of complete case analysis (aka list-wise deletion) in all simulated conditions. We know the theoretical properties of complete case analysis, which makes the technique useful as a point of departure when evaluating multiple imputation performance. 
The evaluation criteria may depend on the estimand. When descriptive statistics are the goal and when statistical inference would not be of interest, bias of the estimates would still apply, but standard errors are generally ignored. Chances are that for those who focus on descriptive statistical applications, multiple imputation would not be the mode of choice. In general, I would say that each multiple imputation routine should be evaluated on at least the following points:

@@

>>>>

---
title: "Towards a standardized evaluation of imputation routines"
author: "Hanne Oberman and Gerko Vink"
date: "19-1-2022"
abstract: "Developing new imputation methodology has become a very active field. Unfortunately, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. In this paper we propose a move towards a standardized evaluation of imputation methods. To demonstrate the need for standardization, we highlight a set of potential pitfalls that bring forth a chain of potential problems in the objective assessment of the performance of imputation routines. This may lead to suboptimal use of multiple imputation in practice. Additionally, we suggest a course of action for simulating and evaluating missing data problems."
bibliography: StandardizedEvaluation.bib  
output: 
  pdf_document:
    keep_tex: true
editor_options: 
  chunk_output_type: inline
---

@@

@@

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policy making, demonstrated e.g. by the detailed manual created by the National Research Council [@little2012prevention]. Although a top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers are embracing imputation techniques. After all, the principle of imputation is very intuitive.

The idea behind imputation is to impute (fill in) missing values, to obtain a valid estimate of what could have been. A popular variant is multiple imputation [@rubi76], wherein missing values are imputed multiple times. The data that are thus obtained can be analyzed by standard techniques and, in the case of multiple imputation, the analysis results can be combined into a single inference [using Rubin's rules; @rubi87]. In contrast to ad hoc methods for dealing with missing values (e.g. list-wise deletion, mean imputation, regression imputation, last observation carried forward, indicator method), multiple imputation properly takes the sources of uncertainty that are related to the missingness problem into account.

<!-- [It has been empirically demonstrated time and again that the inferences on multiply imputed data yield unbiased and confidence valid results (through Rubin's rules). But how do we know that this is the case, if we do not have access to the true but unobserved values of the missing entries? And how can we be sure that extensions to the methodology yield valid inferences too? Especially in the field of ML/AI, there are promising new imputation methods, which may yield even sharper imputations than now-standard (semi-)parametric imputation methods. That's why we use simulation studies for the evaluation of imputation routines.] -->

The quality of a solution obtained by imputation depends on the statistical properties of the incomplete data and the degree to which an imputation procedure is able to capture these properties when modeling missing values. In general it holds that modeling missing data becomes more challenging when the amount of missingness increases. However, when (strong) relations in the data are present, the observed parts can hold great predictive power for the models that estimate the missingness. In that case, (multiple) imputation would be substantially more efficient than the ubiquitous list-wise deletion.

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. In such studies a complete dataset is usually generated from a statistical model, another model is used to induce missingness, [then several missing data methods are applied] and a set of evaluation criteria is postulated to evaluate the performance of the imputation methods. However, no golden standard has been established to evaluate these imputation routines and, as a result, the validity of the simulation procedures may differ tremendously from one developer to another. [Nowadays there are a lot of imp alg being propagated, esp ai/ml, whose perf often times seems quite promising. But may not be a fair comparison because of the simulation set-up/dev in ai/ml so fast that steps may be skipped]

@@

To evaluate the ability of an imputation routine to handle missingness, a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand, by choosing a data generating mechanism to obtain a complete dataset. Complete data is typically either generated from a statistical model, or sampled from a sufficiently large observed set [model-based vs design-based simulation].

@@

<!-- als je perf based sims doet, dan kijk je wel daarnaar. -->

@@

[Briefly introduce missingness mechanisms here and show potential problems, but without simulation results (moved to separate section). This part should contain: 1) definitions of mechanisms, 2) that MCAR is often ignored, while it's the minimum requirement and reference, and sometimes realistic in practice, 3) that MAR can be generated *spuriously* if the correlation between variables is low, and 4) that MNAR is often ignored, while it may be the most realistic in practice. And finally something about types of M(N)AR, but what are specific problems with that? ref dance paper rianne]

@@

@@

<!-- als inf het doel is, dan moet je de varianties juist krijgen dus kom je niet weg met m = 1 (want geen onderscheid tussen obs en imp) -->

@@

@@

## Step 3. Evaluate performance

It is wise to evaluate the performance of complete case analysis (aka list-wise deletion) in all simulated conditions. We know the theoretical properties of complete case analysis, which makes the technique useful as a point of departure when evaluating multiple imputation performance. 
The evaluation criteria may depend on the estimand. When descriptive statistics are the goal and when statistical inference would not be of interest, bias of the estimates would still apply, but standard errors are generally ignored. Chances are that for those who focus on descriptive statistical applications, multiple imputation would not be the mode of choice. In general, I would say that each multiple imputation routine should be evaluated on at least the following points:

@@

## Step 4. Report

<!-- [Checklist voor iedere sim studie met alle parameters die we bespreken] -->

<!-- - doel -->

<!-- - 11 punten van stef -->

<!-- - andere punten uit dit paper -->

<!-- - welke variables/features vormen imp model en welke analysis model (congenial, meng 94) -->

<!-- - cross validation? -->

<!-- als pfd met check boxes? -->

Report at least the following:

1. set-up

- aim
- number of simulations

2. data generation

- model vs design based
- data types
- sample vs population (vs bootstrapping??)
- number of observations
- number of variables
- coherence between variables

3. missingness generation

- monotone/univariate vs multivariate
- missingness mechanism
- type of M(N)AR
- missingness proportion

4. missing data methods

- imputation method
- imputation model
- number of imputations
- number of iterations

5. performance evaluation

- analysis model
- estimand
- diagnostic
- ppc

### 2022-03-22 Manuscript/StandardizedEvaluation.tex
Manuscript after discussion on suggestions Hanne, including first version of the checklist.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Towards a standardized evaluation of multiple imputation routines},
  pdfauthor={Hanne Oberman and Gerko Vink},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\usepackage[normalem]{ulem}
% Avoid problems with \sout in headers with hyperref
\pdfstringdefDisableCommands{\renewcommand{\sout}{}}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Towards a standardized evaluation of multiple imputation
routines}
\author{Hanne Oberman and Gerko Vink}
\date{19-1-2022}

@@

Multiple imputation (Donald B. Rubin 1987) is a state-of-the-art
technique for drawing valid conclusions from incomplete data. The
technique has earned a permanent spot in research and policy making,
demonstrated e.g.~by the detailed manual created by the National
Research Council (Roderick J. Little et al. 2012). Although a top-down
enforcement of valid ways to handle missing data is not yet very
pronounced, an increasing amount of researchers are embracing multiple
imputation techniques. After all, the principle of multiple imputation
is very intuitive.

The idea behind multiple imputation is to impute (fill in) the missing
values in incomplete data multiple times, to obtain a valid estimate of
what could have been observed. The multiple datasets that are thus
obtained can be analyzed by standard techniques and the analysis results
can be combined into a single inference. In contrast to ad hoc methods
for dealing with missing values (e.g.~list-wise deletion, mean
imputation, last observation carried forward), multiple imputation
properly takes the sources of uncertainty that are related to the
missingness problem into account. {[}It has been empirically
demonstrated time and again that the inferences on multiply imputed data
yield unbiased and confidence valid results (through Rubin's rules). But
how do we know that this is the case, if we do not have access to the
true but unobserved values of the missing entries? And how can we be
sure that extensions to the methodology yield valid inferences too?
Especially in the field of ML/AI, there are promising new imputation
methods, which may yield even sharper imputations than now-standard
(semi-)parametric imputation methods. That's why we use simulation
studies for the evaluation of imputation routines.{]}

\sout{The quality of a solution obtained by multiple imputation depends
on the statistical properties of the incomplete data and the degree to
which an imputation procedure is able to capture these properties when
modeling missing values. In general it holds that modeling missing data
becomes more challenging when the amount of missingness increases.
However, when (strong) relations in the data are present, the observed
parts can hold great predictive power for the models that estimate the
missingness. In that case, multiple imputation would be substantially
more efficient than the ubiquitous list-wise deletion.}

When evaluating the statistical properties (and thereby the practical
applicability) of imputation methodology, researchers most often make
use of simulation studies. In such studies a complete dataset is usually
generated from a statistical model, another model is used to induce
missingness, {[}then several missing data methods are applied{]} and a
set of evaluation criteria is postulated to evaluate the performance of
the imputation methods. However, no golden standard has been established
to evaluate these imputation routines and, as a result, the validity of
the simulation procedures may differ tremendously from one developer to
another.

@@

To evaluate the ability of an imputation routine to handle \sout{the
non-response model} {[}missing data entries{]}, a form of truth has to
be established. Those who perform simulation studies are in the luxury
position to establish the truth beforehand, by choosing a data
generating mechanism to obtain a complete dataset. Complete data is
typically either generated from a statistical model, or sampled from a
sufficiently large \sout{observed} set {[}model-based vs design-based
simulation{]}. \sout{However, when generating data, two easily
overlooked problems may arise.}

@@

{[}Briefly introduce missingness mechanisms here and show potential
problems, but without simulation results (moved to separate section).
This part should contain: 1) definitions of mechanisms, 2) that MCAR is
often ignored, while it's the minimum requirement and sometimes
realistic in practice, 3) that MAR can be generated \emph{spuriously} if
the correlation between variables is low, and 4) that MNAR is often
ignored, while it may be the most realistic in practice. And finally
something about types of M(N)AR, but what are specific problems with
that?{]}

@@

\sout{However, because multiple imputation was designed as a mode for
statistical inference, a minimal set of statistical properties should at
least be evaluated. We highlight four aspects of evaluation that should
be considered when evaluation imputation routines.}

@@

\hypertarget{illustration}{%
\section{Illustration}\label{illustration}}

{[}Add simulation study + results here!{]}

@@

@@

\begin{verbatim}
## R version 4.1.3 (2022-03-10)
## Platform: aarch64-apple-darwin20 (64-bit)
## Running under: macOS Monterey 12.3
## 
## Matrix products: default
## BLAS:   /Library/Frameworks/R.framework/Versions/4.1-arm64/Resources/lib/libRblas.0.dylib
## LAPACK: /Library/Frameworks/R.framework/Versions/4.1-arm64/Resources/lib/libRlapack.dylib
## 
## locale:
## [1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8
## 
## attached base packages:
## [1] stats     graphics  grDevices utils     datasets  methods   base     
## 
## other attached packages:
## [1] dplyr_1.0.8
## 
## loaded via a namespace (and not attached):
##  [1] rstudioapi_0.13  knitr_1.37       magrittr_2.0.2   tidyselect_1.1.2
##  [5] R6_2.5.1         rlang_1.0.2      fastmap_1.1.0    fansi_1.0.2     
##  [9] highr_0.9        stringr_1.4.0    tools_4.1.3      xfun_0.30       
## [13] utf8_1.2.2       DBI_1.1.2        cli_3.2.0        htmltools_0.5.2 
## [17] ellipsis_0.3.2   assertthat_0.2.1 yaml_2.3.5       digest_0.6.29   
## [21] tibble_3.1.6     lifecycle_1.0.1  crayon_1.5.0     purrr_0.3.4     
## [25] vctrs_0.3.8      glue_1.6.2       evaluate_0.15    rmarkdown_2.13  
## [29] stringi_1.7.6    compiler_4.1.3   pillar_1.7.0     generics_0.1.2  
## [33] pkgconfig_2.0.3
\end{verbatim}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Towards a standardized evaluation of imputation routines},
  pdfauthor={Hanne Oberman and Gerko Vink},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Towards a standardized evaluation of imputation routines}
\author{Hanne Oberman and Gerko Vink}
\date{19-1-2022}

@@

Imputation is a state-of-the-art technique for drawing valid conclusions
from incomplete data. The technique has earned a permanent spot in
research and policy making, demonstrated e.g.~by the detailed manual
created by the National Research Council (Roderick J. Little et al.
2012). Although a top-down enforcement of valid ways to handle missing
data is not yet very pronounced, an increasing amount of researchers are
embracing imputation techniques. After all, the principle of imputation
is very intuitive.

The idea behind imputation is to impute (fill in) missing values, to
obtain a valid estimate of what could have been. A popular variant is
multiple imputation (Donald B. Rubin 1976), wherein missing values are
imputed multiple times. The data that are thus obtained can be analyzed
by standard techniques and, in the case of multiple imputation, the
analysis results can be combined into a single inference {[}using
Rubin's rules; Donald B. Rubin (1987){]}. In contrast to ad hoc methods
for dealing with missing values (e.g.~list-wise deletion, mean
imputation, regression imputation, last observation carried forward,
indicator method), multiple imputation properly takes the sources of
uncertainty that are related to the missingness problem into account.

The quality of a solution obtained by imputation depends on the
statistical properties of the incomplete data and the degree to which an
imputation procedure is able to capture these properties when modeling
missing values. In general it holds that modeling missing data becomes
more challenging when the amount of missingness increases. However, when
(strong) relations in the data are present, the observed parts can hold
great predictive power for the models that estimate the missingness. In
that case, (multiple) imputation would be substantially more efficient
than the ubiquitous list-wise deletion.

When evaluating the statistical properties (and thereby the practical
applicability) of imputation methodology, researchers most often make
use of simulation studies. In such studies a complete dataset is usually
generated from a statistical model, another model is used to induce
missingness, {[}then several missing data methods are applied{]} and a
set of evaluation criteria is postulated to evaluate the performance of
the imputation methods. However, no golden standard has been established
to evaluate these imputation routines and, as a result, the validity of
the simulation procedures may differ tremendously from one developer to
another. {[}Nowadays there are a lot of imp alg being propagated, esp
ai/ml, whose perf often times seems quite promising. But may not be a
fair comparison because of the simulation set-up/dev in ai/ml so fast
that steps may be skipped{]}

@@

To evaluate the ability of an imputation routine to handle missingness,
a form of truth has to be established. Those who perform simulation
studies are in the luxury position to establish the truth beforehand, by
choosing a data generating mechanism to obtain a complete dataset.
Complete data is typically either generated from a statistical model, or
sampled from a sufficiently large observed set {[}model-based vs
design-based simulation{]}.

@@

{[}Briefly introduce missingness mechanisms here and show potential
problems, but without simulation results (moved to separate section).
This part should contain: 1) definitions of mechanisms, 2) that MCAR is
often ignored, while it's the minimum requirement and reference, and
sometimes realistic in practice, 3) that MAR can be generated
\emph{spuriously} if the correlation between variables is low, and 4)
that MNAR is often ignored, while it may be the most realistic in
practice. And finally something about types of M(N)AR, but what are
specific problems with that? ref dance paper rianne{]}

@@

@@

@@

\hypertarget{step-4.-report}{%
\subsection{Step 4. Report}\label{step-4.-report}}

Report at least the following:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  set-up
\end{enumerate}

\begin{itemize}
\tightlist
\item
  aim
\item
  number of simulations
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  data generation
\end{enumerate}

\begin{itemize}
\tightlist
\item
  model vs design based
\item
  data types
\item
  sample vs population (vs bootstrapping??)
\item
  number of observations
\item
  number of variables
\item
  coherence between variables
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  missingness generation
\end{enumerate}

\begin{itemize}
\tightlist
\item
  monotone/univariate vs multivariate
\item
  missingness mechanism
\item
  type of M(N)AR
\item
  missingness proportion
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  missing data methods
\end{enumerate}

\begin{itemize}
\tightlist
\item
  imputation method
\item
  imputation model
\item
  number of imputations
\item
  number of iterations
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  performance evaluation
\end{enumerate}

\begin{itemize}
\tightlist
\item
  analysis model
\item
  estimand
\item
  diagnostic
\item
  ppc
\end{itemize}

@@

\begin{verbatim}
## R version 4.1.2 (2021-11-01)
## Platform: x86_64-w64-mingw32/x64 (64-bit)
## Running under: Windows 10 x64 (build 19042)
## 
## Matrix products: default
## 
## locale:
## [1] LC_COLLATE=Dutch_Netherlands.1252  LC_CTYPE=Dutch_Netherlands.1252   
## [3] LC_MONETARY=Dutch_Netherlands.1252 LC_NUMERIC=C                      
## [5] LC_TIME=Dutch_Netherlands.1252    
## 
## attached base packages:
## [1] stats     graphics  grDevices utils     datasets  methods   base     
## 
## other attached packages:
## [1] dplyr_1.0.8
## 
## loaded via a namespace (and not attached):
##  [1] rstudioapi_0.13  knitr_1.37       magrittr_2.0.2   tidyselect_1.1.2
##  [5] R6_2.5.1         rlang_1.0.2      fastmap_1.1.0    fansi_1.0.2     
##  [9] highr_0.9        stringr_1.4.0    tools_4.1.2      xfun_0.30       
## [13] utf8_1.2.2       DBI_1.1.2        cli_3.2.0        htmltools_0.5.2 
## [17] ellipsis_0.3.2   assertthat_0.2.1 yaml_2.3.5       digest_0.6.29   
## [21] tibble_3.1.6     lifecycle_1.0.1  crayon_1.5.0     purrr_0.3.4     
## [25] vctrs_0.3.8      glue_1.6.2       evaluate_0.15    rmarkdown_2.13  
## [29] stringi_1.7.6    compiler_4.1.2   pillar_1.7.0     generics_0.1.2  
## [33] pkgconfig_2.0.3
\end{verbatim}
